---
title: "Methodology"
description: |
  Step-by-step description of analysis
output: 
  distill::distill_article:
    toc: true
    toc_float: true
    toc_depth: 3
    self_contained: false
---
<style>

body{
  background-color: #EBEDEF;
  font-family:helvetica;
}
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.retina = 2,
                      echo = TRUE,
                      eval = TRUE,
                      message = FALSE,
                      warning = FALSE)
```

## Methodology and Design

In this section, we will explore the methodology to analyze the data and analyze various packages required to build the plots. The criteria for selection of plots are as follows:
    - Level of customization 
    - Ease of use and implementation of customization 
    - Ease of understanding and interpretation of the plot- both clarity and aesthetic
    - Interactivity

### Methodology

#### Exploratory Visualization

1) As there are no common identifiers between the loyalty card and credit card dataset, we will proceed to analyze them separately.

2) To identify the most popular locations and when they are popular, we will analyze the transaction volume based on:
    - day (across the 2 weeks)
    - day of week

3) Time information is provided for credit card transactions but not loyalty card transactions. Nonetheless, we will extract the date from datetime information provided in credit card dataset so that we can analyze both datasets across the common dates.

```{r}
glimpse(loyalty_data)
```

```{r}
glimpse(cc_data)
```

#### GeoVisual Analysis

1) **Georeferencing**

Given that we are provided with the tourist map and SHP file of Abila, we are able to perform georeferencing. We will perform georeferencing in QGIS.

2) **Drawing the paths** 

We are provided with longitude and latitude of each car across the 2 weeks, allowing us to plot their trajectories and hence locate each car and its driver at a certain point of a day. 

3) **Understand the paths**

After plotting the trajectory of each car on the tourist map in step 2, we are able to visualize where each driver was at at each time of the day. As the data provided does not match the credit and loyalty card to its owner, we can use the GPS trajectories to match against the credit card information based on location and timestamp to match the driver to the card.

4) **Interpretations on "patterns of life" analysis**

For the daily patterns, we look at the activities of different vehicles at different hours of the day – what time they go to the office, what time they leave and where do they go at different hours. For the visualization design, an interactive map is used to allow the users to utilize the tooltip function to get information on time, to be used for the analysis easily.

#### Network Visualisation and Analysis

1) **Segmenting employees**

The employees were divided into their respective departments. This sets the scene for the interpretation of the connections in the subsequent sessions, as the patterns will be analyzed for each segment respectively with the certain patterns being compared between similar segments.

2) **Drawing the network graph**

Bipartite graphs are used as we are provided with credit card information showing employees transactions with each location. As there is no information on interactions between employees themselves, we will utilize a bipartite graph where two sets of vertices which are connected to each other, but not within themselves.

3) **Interpreting the connections**

Due to the limitations in RMarkdown, crosstalk is not available for the VisNetwork package used to visualize networks. The network graph shows connections between an employee and the location he/she visits based on the credit card transaction data provided. However, it does not reflect when these visits occur and if various employees may have crossed paths due to transactions at the same location during the same time period. Hence we will utilize both the interactive network graph and an interactive datatable to get further details on these transactions such as the time they occurred to assess if the employees may have crossed paths.

#### Overall visualization design concepts

The visualizations were designed make full use of RStudio's interactive plots such that we can provide a tool for the users to customize their analysis as much as possible. The users are able to filter the different input parameters (such as location for exploratory visualization and by department for network graph) and utilize the maps tooltips function and interactive datatables to obtain greater detail on the visualizations. This will be explained in greater details in the plot design section.

### Plot Design

#### Exploratory Visualization

1) **Transaction volume overview by day**- We have decided to employ the use of a linked bar chart to reflect the volume of transactions across the 2 weeks for each location. Users can then select the location they want specific details on and the linked line chart will then provide the details on transaction volume across time.

```{r, tidy.opts=list(width.cutoff=55),tidy=TRUE, layout="l-body-outset", fig.width=5.5, fig.height=5, fig.cap="Bar Chart of loyalty card transaction volume per location cross-linked with line chart"}
##Volume from loyalty card dataset by date- linked

loyalty_data_cleaned_grpcount <- loyalty_data_cleaned %>%
  group_by(location, date) %>%
  summarise(txn_count = n())

shared_data <- SharedData$new(loyalty_data_cleaned_grpcount, key = ~location)

loyalty_loc_vol_bar <- shared_data %>%
  plot_ly() %>% 
  group_by(location) %>%
  summarise(txn_count = sum(txn_count)) %>%
  arrange(desc(txn_count)) %>%
  add_bars(x = ~txn_count, y = ~location, type = "bar", hoverinfo = "text", 
          text = ~paste("Location:", location,
                        "<br>Transaction Volume:", txn_count)) %>%
  layout(title = "Loyalty card transaction Volume by Location", xaxis = list(title = "Transaction Volume", 
         zeroline = FALSE), yaxis = list(title = "Location", zeroline = FALSE, categoryorder = "array", dtick= 1, categoryarray = ~txn_count))

loyalty_loc_vol_line <- shared_data %>% 
  plot_ly() %>%
  add_lines(x = ~date, y = ~txn_count, hoverinfo = "text", 
          text = ~paste("Location:", location,
                        "<br>Date:", date,
                        "<br>Transaction Volume:", txn_count)) %>%
  layout(xaxis = list(title = "Date", showgrid = TRUE, showticklabels = TRUE), 
         yaxis = list(title = "Transaction volume"))

bscols(widths = c(12,12), loyalty_loc_vol_bar, loyalty_loc_vol_line)
```

```{r, tidy.opts=list(width.cutoff=55),tidy=TRUE, layout="l-body-outset", fig.width=5.5, fig.height=5, fig.cap="Bar Chart of credit card transaction volume per location cross-linked with line chart"}
##Volume from credit card dataset by date- linked

cc_data_cleaned_grpcount <- cc_data_cleaned %>%
  group_by(location, date) %>%
  summarise(txn_count = n())

shared_data_cc <- SharedData$new(cc_data_cleaned_grpcount, key = ~location)

cc_data_loc_vol_bar <- shared_data_cc %>%
  plot_ly() %>% 
  group_by(location) %>%
  summarise(txn_count = sum(txn_count)) %>%
  arrange(desc(txn_count)) %>%
  add_bars(x = ~txn_count, y = ~location, type = "bar", hoverinfo = "text", 
          text = ~paste("Location:", location,
                        "<br>Transaction Volume:", txn_count)) %>%
  layout(title = "Credit card transaction Volume by Location", xaxis = list(title = "Transaction Volume", 
         zeroline = FALSE), yaxis = list(title = "Location", zeroline = FALSE, categoryorder = "array", dtick= 1, categoryarray = ~txn_count))

cc_loc_vol_line <- shared_data_cc %>% 
  plot_ly() %>%
  add_lines(x = ~date, y = ~txn_count, hoverinfo = "text", 
          text = ~paste("Location:", location,
                        "<br>Date:", date,
                        "<br>Transaction Volume:", txn_count)) %>%
  layout(xaxis = list(title = "Date", showgrid = TRUE, showticklabels = TRUE), 
         yaxis = list(title = "Transaction volume"))

bscols(widths = c(12,12), cc_data_loc_vol_bar, cc_loc_vol_line)
```

2) **Overview of transaction volume difference**- provide an overview of the difference in transaction volume across each location per day.

```{r, tidy.opts=list(width.cutoff=55),tidy=TRUE, layout="l-body-outset", fig.width=7.5, fig.height=5.5, fig.cap="Heatmap showing difference in transaction volume per location"}
##Difference in volume by day

loyalty_data_count <- loyalty_data_cleaned %>%
  group_by(location, date) %>%
  summarise(txn_count = n()) %>%
  arrange(desc(txn_count))

cc_data_count <- cc_data_cleaned %>%
  group_by(location, date) %>%
  summarise(txn_count = n()) %>%
  arrange(desc(txn_count))

count_loyalty_cc_comb <- loyalty_data_count %>%
  full_join(cc_data_count, by = c("location","date")) %>%
  rename("loyalty_card_txncount" = "txn_count.x", "credit_card_txncount" = "txn_count.y") %>%
  mutate_if(is.numeric , replace_na, replace = 0) %>%
  mutate(difference_vol = loyalty_card_txncount - credit_card_txncount) 

a <- list(
  title = "Difference in volume per day by location",
  showticklabels = TRUE,
  dtick = 1
)

txn_vol_difference_heatmap <- count_loyalty_cc_comb %>%
  plot_ly(x= ~date, y = ~location , z = ~difference_vol, colors = brewer.pal(3, "Blues"), type = "heatmap",   hoverinfo = 'text',
          text = ~paste("Date:", date,
                        "<br> Location:", location,
                        "<br> Vol_Diference:", difference_vol))
txn_vol_difference_heatmap %>% layout(yaxis = a, xaxis = list(
  dtick = "86400000.0",
  type = 'date',
  title = "Date",
  tickangle = 0,
  tickformat = "%d%b"), margin = m, plot_bgcolor="#bdbdbd")
```

3) **Transaction volume overview by day of week**- Similar to 1 above, We have decided to employ the use of a linked bar chart to reflect the volume of transactions by each day of the week for each location. Users can use the interactive filter to select the location they want specific details on and the linked bar chart will then provide the visualization on transaction volume across day of week. The datatable will also be filtered for the relevant location to provide further details like date and price of transaction.

As the transaction time for credit cards is also provided, we have included an additional bar chart which reflects the volume per period of time in the selected location. This will provide more information to assess when the location is popular for each day of the week.

```{r, tidy.opts=list(width.cutoff=60),tidy=TRUE, fig.cap="Bar Chart of loyalty card transaction volume per location by day of week cross-linked with datatable"}
##Loyalty card volume by day of week

##Preparing the data
loyalty_data_cleaned$daynumber = lubridate::wday(loyalty_data_cleaned$date, week_start = 1)
loyalty_data_cleaned$weekday = factor(loyalty_data_cleaned$weekday, levels= c("Sunday", "Monday", 
                                             "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday"))



##Plots
shared_data_loyalty = SharedData$new(loyalty_data_cleaned, key = ~ location)

loyalty_wkday_vol <- shared_data_loyalty %>%
  plot_ly(width=700, height=500) %>%
  group_by(weekday,location) %>%
  summarize(transaction = n()) %>%
  arrange(desc(transaction)) %>%
  add_trace(x = ~weekday, y = ~transaction, color = ~location, type = 'bar',hoverinfo = 'text',
            hovertext  = ~paste("<br> Day_of_week:", weekday,
                          "<br> Transaction Volume:", transaction)) %>%
  layout(xaxis = list(title = "Day_of_week", 
                        type = "factor",
                        categoryorder = "factor",
                        categoryarray = loyalty_data_cleaned$weekday[order(loyalty_data_cleaned[['daynumber']])],
                        showgrid = TRUE,
                        showticklabels = TRUE
                      ), yaxis = list(title = "Transaction volume"))

bscols(widths = c(10,4),
  list(
    filter_select("location", "Please Specify the location", shared_data_loyalty, group =  ~location, multiple = F),
    datatable(shared_data_loyalty) %>% formatDate(1, method = 'toLocaleDateString')),
  loyalty_wkday_vol
)
```

```{r, tidy.opts=list(width.cutoff=55),tidy=TRUE, fig.cap="Bar Chart of credit card transaction volume per location by day of week and timegroup, cross-linked with datatable"}
## Credit card volume by day of week
## Loading data
Sys.setlocale("LC_TIME", "English")
## Preparing the data
cc_data_cleaned$daynumber = lubridate::wday(cc_data_cleaned$timestamp, week_start = 1)
cc_data_cleaned$weekday = factor(cc_data_cleaned$weekday, levels= c("Sunday", "Monday", 
                                             "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday"))

## Preparing timegroup
cc_data_cleaned$timegroup = ""
cc_data_cleaned$timegroup[cc_data_cleaned$hour < 05 & 00 <= cc_data_cleaned$hour] = "00 - 04"
cc_data_cleaned$timegroup[cc_data_cleaned$hour < 09 & 05 <= cc_data_cleaned$hour] = "05 - 08"
cc_data_cleaned$timegroup[cc_data_cleaned$hour < 13 & 09 <= cc_data_cleaned$hour] = "09 - 12"
cc_data_cleaned$timegroup[cc_data_cleaned$hour < 17 & 13 <= cc_data_cleaned$hour] = "13 - 16"
cc_data_cleaned$timegroup[cc_data_cleaned$hour < 21 & 17 <= cc_data_cleaned$hour] = "17 - 20"
cc_data_cleaned$timegroup[cc_data_cleaned$hour <= 24 & 21 <= cc_data_cleaned$hour] = "21 - 24"

## Plots
shared_data = SharedData$new(cc_data_cleaned, key = ~ location)

cc_wkday_vol <- shared_data %>%
  plot_ly(width=700, height=500) %>%
  group_by(weekday,location) %>%
  summarize(transaction = n()) %>%
  arrange(desc(transaction)) %>%
  add_trace(x = ~weekday, y = ~transaction, color = ~location, type = 'bar',hoverinfo = 'text',
            hovertext  = ~paste("<br> Day_of_week:", weekday,
                          "<br> Transaction Volume:", transaction)) %>%
  layout(xaxis = list(title = "Day_of_week", 
                        type = "factor",
                        categoryorder = "factor",
                        categoryarray = cc_data_cleaned$weekday[order(cc_data_cleaned[['daynumber']])],
                        showgrid = TRUE,
                        showticklabels = TRUE), 
                        yaxis = list(title = "Transaction volume"))

cc_timegroup_vol <-shared_data %>%
  plot_ly( width=700, height=500) %>%
  group_by(weekday,timegroup,location) %>%
  summarize("transaction" = n()) %>%
  add_trace(x = ~weekday, y = ~transaction, color = ~timegroup, type = 'bar',hoverinfo = 'text',
            hovertext  = ~paste("<br> Time:", timegroup,
                                "<br>Transaction Volume:", transaction)) %>%
  layout(xaxis = list(title = "Time",  type = "factor",
                      categoryorder = "factor",
                      categoryarray = cc_data_cleaned$weekday[order(cc_data_cleaned[['daynumber']])],
                      showgrid = TRUE,
                      showticklabels = TRUE), 
                      yaxis = list(title = "Transaction volume"))



bscols(widths = c(10,4),
  list(
    filter_select("location", "Please Specify the location", shared_data, group =  ~location, multiple = F),
  cc_wkday_vol,
  cc_timegroup_vol, 
  datatable(shared_data) %>% formatDate(1, method = 'toLocaleDateString')
))
```

#### GeoVisual Analysis

1) **Trajectory analysis**- To visualize the path traveled by each car. We have employed both the use of an interactive map and static facet map.
    - Interactive map- Clicking on any point in the trajectory allows us to see the CarID, day and timestamp of the respective point in the route.
    - Facet map- This allows us to visualize a single variable with many levels and arrange the plots in a more space efficient manner for easy comparison and analysis.

```{r, tidy.opts=list(width.cutoff=60),tidy=TRUE, fig.cap="Interactive map"}
##Plotting GPS paths
gps_path_selected <- gps_path_cleaned %>%
  filter(id == 1, day == 6)

tmap_mode("view")
tm_shape(bgmap) +
  tm_rgb(bgmap, r = 1, g = 2, b = 3, alpha = NA, saturation = 1, interpolate = TRUE, max.value = 255) +
  tm_shape(gps_path_selected) +
  tm_lines(col= "red")
```

```{r, tidy.opts=list(width.cutoff=60),tidy=TRUE, layout="l-body-outset", fig.width=6, fig.height=20, fig.cap="Facet map"}
##Facet map
gps_path_selected <- gps_path_cleaned %>%
  filter(hour == 12)
  
tmap_mode("plot")
tm_shape(bgmap) +
  tm_rgb(bgmap, r = 1, g = 2, b = 3, alpha = NA, saturation = 1, interpolate = TRUE, max.value = 255) +
  tm_shape(gps_path_selected) +
  tm_lines() +
  tm_facets(by = "id", ncol = 3)
```

#### Network Visualisation and Analysis

1) **Network analysis**- To visualize the possible associations between employees by analyzing their transaction history. Bipartite graphs are used as we are provided with credit card information showing employees transactions with each location. As there is no information on interactions between employees themselves, we will utilize a bipartite graph where two sets of vertices which are connected to each other, but not within themselves.

We have employed both the use of an interactive and static visualization.
    - Interactive network graph- This allows the user to select any node on the graph and observe the interactions between this node and the other nodes will be highlighted, while the other nodes will be dulled. The interactive filtering option also allows user to select by department and id. In order to assess if the employees have crossed paths during their visits to certain locations, we will use the interactive network graph concurrently with an interactive datatable where filters have been enabled to allow filtering for the respective location.
    
    - Facet network graph- This allows us to visualize a single variable with many levels and arrange the plots in a more space efficient manner for easy comparison and analysis. In this case, we have presented a facet graph by day and timegroup, allowing users to observe the interactions between nodes for each timegroup in day 6.

```{r, layout="l-body-outset", fig.width=8, fig.height=7, fig.cap="Interactivet network graph"}
##Interactive viz
visNetwork(cc_nodes, cc_edges, height = "800px", width = "100%") %>%
  visIgraphLayout(layout = "layout_in_circle") %>%
  visOptions(highlightNearest = TRUE, nodesIdSelection = TRUE, selectedBy = "group") %>%
  visLegend() %>%
  visGroups(groupname = "Security", color = "#FFFFA3") %>%
  visGroups(groupname = "Engineering", color = "#FFAFB1") %>%
  visGroups(groupname = "Information Technology", color = "#A1EC76") %>%
  visGroups(groupname = "Facilities", color = "#F0B3F5") %>%
  visGroups(groupname = "Executive", color = "#FF3333") %>%
  visLayout(randomSeed = 123)
```

```{r, layout="l-body-outset", fig.width=12, fig.height=10, fig.cap="Facet map of day by timegroup"}
##Facet
cc_edges <- cc_edges %>%
  filter(day == 6)

cc_graph <- tbl_graph(nodes = cc_nodes,
                      edges = cc_edges,
                      directed = FALSE)

set_graph_style()
g <- ggraph(cc_graph, layout = "nicely") + 
  geom_edge_link(aes(width = weight), alpha=0.2) +
  scale_edge_width(range = c(0.1, 5)) +
  geom_node_point(aes(colour = group), size = 2) 

g + facet_edges(day~timegroup) +
th_foreground(foreground = "grey80",
border = TRUE) +
theme(legend.position = 'bottom')
```
